---
title: "Estimating 6-year lung cancer risk with PLCOm2012"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating 6-year lung cancer risk with PLCOm2012}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(PLCOm2012)
```

## What this package does

`PLCOm2012` implements the lung cancer risk model of Tammemägi et al.
(*NEJM* 2013). Given an ever-smoker's age, race/ethnicity, education, BMI,
COPD status, personal and family cancer history, and smoking history, it returns
the **6-year probability of a first lung cancer**.

There are two ways to use it:

* `plcom2012()` — the vectorised model function, for analysis in R.
* `model_run()` / `get_sample_input()` / `get_default_input()` — the
  ModelsCloud prediction API, for hosting the model as a service.

## Scoring one person

```{r}
plcom2012(
  age = 62, race = "White", education = 4, bmi = 27,
  copd = 0, cancer_hist = 0, family_hist_lung_cancer = 0,
  smoking_status = 0, smoking_intensity = 20,
  duration_smoking = 27, smoking_quit_time = 10
)
```

Heavier exposure and comorbidity raise the estimate:

```{r}
plcom2012(
  age = 70, race = "White", education = 2, bmi = 24,
  copd = 1, cancer_hist = 0, family_hist_lung_cancer = 1,
  smoking_status = 1, smoking_intensity = 40,
  duration_smoking = 45, smoking_quit_time = 0
)
```

## Scoring a cohort

`plcom2012()` is fully vectorised — pass vectors and scalars are recycled:

```{r}
cohort <- data.frame(
  age               = c(58, 64, 72),
  race              = c("White", "Black", "Asian"),
  education         = c(5, 2, 4),
  bmi               = c(22, 28, 25),
  copd              = c(0, 1, 0),
  cancer_hist       = c(0, 0, 1),
  family_hist       = c(0, 1, 0),
  smoking_status    = c(0, 1, 0),
  smoking_intensity = c(15, 35, 20),
  duration_smoking  = c(30, 44, 35),
  quit_time         = c(8, 0, 12)
)

cohort$risk <- plcom2012(
  age = cohort$age, race = cohort$race, education = cohort$education,
  bmi = cohort$bmi, copd = cohort$copd, cancer_hist = cohort$cancer_hist,
  family_hist_lung_cancer = cohort$family_hist,
  smoking_status = cohort$smoking_status,
  smoking_intensity = cohort$smoking_intensity,
  duration_smoking = cohort$duration_smoking,
  smoking_quit_time = cohort$quit_time
)
cohort[, c("age", "race", "risk")]
```

## Using the ModelsCloud API

The same package exposes the standard ModelsCloud prediction interface. Start
from a worked example or the default baseline patient and modify it:

```{r}
get_default_input()

patient <- get_default_input()
patient$age  <- 68
patient$copd <- 1
model_run(patient)
```

`model_run()` returns the input augmented with `risk` (probability) and
`risk_percent`:

```{r}
model_run(get_sample_input())
```

## Race / ethnicity coding

`race` accepts case-insensitive labels (with common aliases) or numeric codes:

| Code | Category | Aliases |
|--|--|--|
| 1 | White | Caucasian, Non-Hispanic White |
| 2 | Black | African American |
| 3 | Hispanic | Latino, Latina |
| 4 | Asian | |
| 5 | American Indian or Alaskan Native | American Indian, Alaskan Native, Indigenous |
| 6 | Native Hawaiian or Pacific Islander | Native Hawaiian, Pacific Islander, Hawaiian |

```{r}
identical(
  plcom2012(age = 60, race = "African American", education = 3, bmi = 26,
            copd = 0, cancer_hist = 0, family_hist_lung_cancer = 0,
            smoking_status = 1, smoking_intensity = 20,
            duration_smoking = 30, smoking_quit_time = 0),
  plcom2012(age = 60, race = 2, education = 3, bmi = 26,
            copd = 0, cancer_hist = 0, family_hist_lung_cancer = 0,
            smoking_status = 1, smoking_intensity = 20,
            duration_smoking = 30, smoking_quit_time = 0)
)
```

## Caveats

* The model is for **ever-smokers**; `smoking_intensity` must be > 0.
* Risk is a probability; eligibility thresholds (e.g. ≥1.5%) are a separate,
  program-specific decision and are intentionally **not** applied here.
* For research use only — not a medical device.

## Reference

Tammemägi MC, Katki HA, Hocking WG, et al. Selection criteria for lung-cancer
screening. *N Engl J Med.* 2013;368(8):728–736.
doi:[10.1056/NEJMoa1211776](https://doi.org/10.1056/NEJMoa1211776)
